---
title: "Assignment 5 - Fall 2024 - MIT 18.642"
format: pdf
editor: visual
---

# 18.642 Assignment 5 Fall 2024

## P1

Suppose ${W (t)}$ is a Brownian motion model with drift $\mu \in R$ and volatility $\sigma$\> 0.

### (b)

Simulate the path of a Brownian Motion Process with the following properties:

$T$ = 1(year)

$\mu$ = 0.30

$\sigma$ = 0.40

$m$ = 252 increments, i.e., $t_i = i × (T/m), i =1, 2,...,m.$

Plot five simulated paths of the process (plotting just the process values at the time increments).

Answer:

We have that $W(t)=\mu t + \sigma B(t)$ where $B(t)$ is the standard Brownian motion.

```{r}
# Set seed for reproducibility
set.seed(10)

# Parameters
T <- 1                  # Time horizon (1 year)
mu <- 0.30              # Drift
sigma <- 0.40           # Volatility
m <- 252                # Number of increments
dt <- T / m             # Time step

# Time grid
time_grid <- seq(0, T, length.out = m + 1)

# Number of simulated paths
n_paths <- 5

# Matrix to store paths (rows = time, columns = paths)
W <- matrix(0, nrow = m + 1, ncol = n_paths)

# Simulate paths
for (j in 1:n_paths) {
  
  # Generate Brownian increments vector
  dB <- rnorm(m, mean = 0, sd = sqrt(dt))
  
  # Construct standard Brownian motion vector
  B <- c(0, cumsum(dB))
  
  # Add drift
  W[, j] <- mu * time_grid + sigma * B
}

# Plot the paths
matplot(time_grid, W, type = "l", lty = 1, lwd = 2,
        xlab = "Time (years)", ylab = "W(t)",
        main = "Simulated Brownian Motion with Drift (5 Paths)")

# Add legend
legend("topleft", legend = paste("Path", 1:n_paths),
       col = 1:n_paths, lty = 1, lwd = 2, cex = 0.7)
```

## P2

The stochastic process {$N(t), t > 0$}, is a Poisson process with intensity or jump-rate $λ$ if:\
- For any $n$ times: $0 = t_0 <t_1 < ··· <t_n$, the increments $N(t_k)−N(t_{k−1})$ are independent, for $1 ≤ k ≤ n$.

\- The distribution of $N(t) − N(s)$ is Poisson with (mean) parameter $λ(t − s)$, for all $0 ≤ s<t< ∞$.

### (b)

Simulate the path of a Poisson Process with the following properties:

$T$ = 1(year)

$\lambda$ = 24 (rate of 24 events per year)

$m$ = 252 increments, i.e., $t_i=(T/m), i = 1, 2,...,m.$

Plot five simulated paths of the process (plotting just the process values at the time increments).

Answer:

```{r}
# Set seed for reproducibility
set.seed(10)

# Parameters
T <- 1                  # Time horizon (1 year)
lambda <- 24            # Rate (events per year)
m <- 252                # Number of increments
dt <- T / m             # Time step

# Time grid
time_grid <- seq(0, T, length.out = m + 1)

# Number of simulated paths
n_paths <- 5

# Matrix to store paths
N <- matrix(0, nrow = m + 1, ncol = n_paths)

# Simulate paths
for (j in 1:n_paths) {
  
  # Generate Poisson increments
  dN <- rpois(m, lambda * dt)
  
  # Construct Poisson process path
  N[, j] <- c(0, cumsum(dN))
}

# Plot the paths
matplot(time_grid, N, type = "s", lty = 1, lwd = 2,
        xlab = "Time (years)", ylab = "N(t)",
        main = "Simulated Poisson Process (5 Paths)")

# Add legend
legend("topleft", legend = paste("Path", 1:n_paths),
       col = 1:n_paths, lty = 1, lwd = 2, cex = 0.7)
```
