---
title: "Assignment 2 - Fall 2024 MIT18642 .qmd"
format: pdf
editor: visual
---

# 18.642 Assignment 2 Fall 2024

## P2

### (c)

Suppose a stock has price $S_0$ at $t = 0$ and at time t (years) it has price
$S_0×X$, where $X$ has lognormal distribution with parameters $µ = t×ln(1.15)$ and $σ = \sqrt{t}×ln(1.3)$. Find the mean and standard deviation of $X$ (the total return) for horizons t = 1/12, 3/12, 1 (years), i.e., corresponding to horizons of 1 month, 3 months and 1 year.

```{r}
# Loop over each time value
for (t in c(1,3,12)){ 
  cat("t=",t,"/12:\n") 
  
  # Parameters of the lognormal distribution
  mu <- t * log(1.15)
  sigma <- sqrt(t) * log(1.3)
  
  # Compute mean and variance of lognormal return
  mean <- exp(mu + 0.5 * sigma^2)
  stdev  <- (mean^2) * (exp(sigma^2) - 1)
  
  # Print results
  cat(paste0("The mean of X for t = ", t, "/12 is ", mean, "\n"))
  cat(paste0("The standard deviation of X for t = ", t, "/12 is ",stdev, "\n"))
  
} 
```

### (e) 

For the stock in part c) if $S_0$ = \$100 compute the expected payoff of the call option when T = 1/12 (1 month), T = 3/12 (3 months) and T = 1 (1 year) for two cases of the strike price:
- K = \$100 (an ‘At-The-Money’ Call)
- K = \$120 (an ‘Out-of-The-Money’ Call)
Comment on the impact of increasing the time to expiration of the call option.

```{r}
S <- 100
# Loop over each time value
for (sp in c(100,120)){
  cat("\nStrike price K=",sp,":\n")
  
  for (t in c(1,3,12)){ 
    cat("t=",t,"/12:\n") 
    
    # Parameters of the lognormal distribution
    mu <- t * log(1.15)
    sigma <- sqrt(t) * log(1.3)
    K<-sp/S
    
    # Compute mean and variance of lognormal return
    mean <- exp(mu + 0.5 * sigma^2)
    p_1 <- sigma + (mu-log(K))/sigma 
    p_2 <- (mu-log(K))/sigma 
    payoff <- S*(mean*pnorm(p_1)-K*pnorm(p_2))
    # Print results
    answer="The expected payoff of the call option when T = "
    cat(paste0(answer, t, "/12 is ", payoff,"\n"))

  }
}
```

Comment on the impact of increasing the time to expiration of the call option.

Answer: An increase in the time to expiration leads to a higher payoff and an increase in the strike price leads to a lower payoff.
