---
title: "Assignment 5 - Fall 2024 - MIT 18.642"
editor: visual
format:
  pdf:
    output-file: "Assignment 5 - Fall 2024 MIT18642 - Solution"
    toc: true
    number-sections: true
  html:
    output-file: "Assignment 5 - Fall 2024 MIT18642 - Solution"
    toc: true
    number-sections: true
    embed-resources: true
    image-format: "png"
---

\newpage

# Problem 1. Brownian Motion with drift and volatility

Suppose ${W (t)}$ is a Brownian motion model with drift $\mu \in R$ and volatility $\sigma$\> 0.

## 1 (b)

Simulate the path of a Brownian Motion Process with the following properties:

$T$ = 1(year)

$\mu$ = 0.30

$\sigma$ = 0.40

$m$ = 252 increments, i.e., $t_i = i × (T/m), i =1, 2,...,m.$

Plot five simulated paths of the process (plotting just the process values at the time increments).

Answer:

We have that $W(t)=\mu t + \sigma B(t)$ where $B(t)$ is the standard Brownian motion.

```{r}
# Set seed for reproducibility
set.seed(10)

# Parameters
T <- 1                  # Time horizon (1 year)
mu <- 0.30              # Drift
sigma <- 0.40           # Volatility
m <- 252                # Number of increments
dt <- T / m             # Time step

# Time grid
time_grid <- seq(0, T, length.out = m + 1)

# Number of simulated paths
n_paths <- 5

# Matrix to store paths (rows = time, columns = paths)
W <- matrix(0, nrow = m + 1, ncol = n_paths)

# Simulate paths
for (j in 1:n_paths) {
  
  # Generate Brownian increments vector
  dB <- rnorm(m, mean = 0, sd = sqrt(dt))
  
  # Construct standard Brownian motion vector
  B <- c(0, cumsum(dB))
  
  # Add drift
  W[, j] <- mu * time_grid + sigma * B
}

# Plot the paths
matplot(time_grid, W, type = "l", lty = 1, lwd = 2,
        xlab = "Time (years)", ylab = "W(t)",
        main = "Simulated Brownian Motion with Drift (5 Paths)")

# Add legend
legend("topleft", legend = paste("Path", 1:n_paths),
       col = 1:n_paths, lty = 1, lwd = 2, cex = 0.7)


```

# Problem 2. Poisson Process

The stochastic process {$N(t), t > 0$}, is a Poisson process with intensity or jump-rate $λ$ if:\
- For any $n$ times: $0 = t_0 <t_1 < ··· <t_n$, the increments $N(t_k)−N(t_{k−1})$ are independent, for $1 ≤ k ≤ n$.

\- The distribution of $N(t) − N(s)$ is Poisson with (mean) parameter $λ(t − s)$, for all $0 ≤ s<t< ∞$.

## 2 (b)

Simulate the path of a Poisson Process with the following properties:

$T$ = 1(year)

$\lambda$ = 24 (rate of 24 events per year)

$m$ = 252 increments, i.e., $t_i=(T/m), i = 1, 2,...,m.$

Plot five simulated paths of the process (plotting just the process values at the time increments).

Answer:

```{r}
# Set seed for reproducibility
set.seed(10)

# Parameters
T <- 1                  # Time horizon (1 year)
lambda <- 24            # Rate (events per year)
m <- 252                # Number of increments
dt <- T / m             # Time step

# Time grid
time_grid <- seq(0, T, length.out = m + 1)

# Number of simulated paths
n_paths <- 5

# Matrix to store paths
N <- matrix(0, nrow = m + 1, ncol = n_paths)

# Simulate paths
for (j in 1:n_paths) {
  
  # Generate Poisson increments
  dN <- rpois(m, lambda * dt)
  
  # Construct Poisson process path
  N[, j] <- c(0, cumsum(dN))
}

# Plot the paths
matplot(time_grid, N, type = "s", lty = 1, lwd = 2,
        xlab = "Time (years)", ylab = "N(t)",
        main = "Simulated Poisson Process (5 Paths)")

# Add legend
legend("topleft", legend = paste("Path", 1:n_paths),
       col = 1:n_paths, lty = 1, lwd = 2, cex = 0.7)
```

# Problem 3. Extreme value method for estimating the variance of the rate of return.

## 3(a)

Read the article Parkinson (1980).

## 3(b)

Explain the definition of the term $D_x$ in Section II.

Answer:

If $X_t$ has a probability density function $f(x)=\frac{1}{\sqrt{2\pi Dt}}exp(\frac{-(x-x_0)^2}{2Dt})$ where $D$ is the diffusion constant. Then we can estimate D using $D_x$: $$
D_x = \frac{1}{n-1} \sum_{i=1}^{n} (d_i - \bar{d})^2
$$ where $d_i=x_i-x_{i-1}$ is the displacement during the ith interval and $\bar{d} = \frac{1}{n} \sum_{i=1}^{n} d_i$ is the sample mean displacement.

## 3(c)

Explain the definition of the term $D_l$ in Section III.

Answer:

$D_l$ is an estimator of $D$ based on the method of moments and the identity $\mathbb{E}(l^2) = 4 \ln(2)\, D t$. Rearranging, we obtain:\
$$D=\frac{1}{4 \ln(2)}\mathbb{E}(l^2) \approx 0.361 \cdot \mathbb{E}(l^2) \quad \text{for } t = 1$$Hence, $D_l$ is obtained by replacing the expectation with its empirical analogue, namely the sample mean of the squared ranges computed over $n$ unit time intervals: $$
D_l =0.361 \  \frac{1}{n} \sum_{i=1}^{n} l_i^2
$$ where $l_i = \max X_t- \min X_t$ denotes the observed range of the process over the $i$-th unit time interval.

## 3(d)

Using the results of Section IV, explain why the improvement using $D_l$, as an estimate of V (in Section V) could be of particular importance in studies of the time and price dependence (if any) of V.

Answer:

The improvement obtained by using $D_l$ instead of $D_x$ to estimate the variance of the rate of return is that less data is required to compute an estimate with the same accuracy.

# Problem 4. Volatility Case Studies

## 4(a)

Detail/comment on specific results of the SP500 case study you consider interesting/significant.

## 4(b)

Detail/comment on specific results of the ARKK case study you consider interesting/significant.

## 4(c)

Compare the results/findings about volatility for ARKK versus the SP500.

# Problem 5. Optional / Extra Credit

## 5(a)

Create the pdf file of the case study.

## 5(b)

Detail/comment on specific results you consider interesting/significant.

## 5(c)

Compare the results/findings for your stock to those for the SP500 and ARKK.
