---
title: "One-Period Financial Models - MIT 18.642"
editor: visual
execute:
  echo: true

format:
  pdf:
    output-file: "mit18_642_f24_lec02_2 - R code One-Period Financial Models"
    toc: true
    number-sections: true
    
  html:
    output-file: "mit18_642_f24_lec02_2 - R code One-Period Financial Models"
    embed-resources: true
    image-format: "png"
    toc: true
    toc-location: "left-body"
    number-sections: true
---

::: {.content-visible when-format="pdf"}
\newpage
:::

# Lecture 2 - MIT 18.642- One-Period Financial Models

## Visualizing Assets/Portfolios with ggplot2

```{r}
# Define the row matrix of t=0 prices B_0 and S_0 
A0=matrix(c(100,100),nrow=1,ncol=2) 
# Add dimension names for rows and columns 
dimnames(A0)<-list(c("Price_0"), c("Bond","Stock")) 
A0
```

```{r}
# Define the matrix of time T payoffs/prices for B_T and S_T 
# where row 1 corresponds to state \omega=d and 
#       row 2 corresponds to state \omega=u 
#       column 1 corresponds to the Bond 
#       column 2 corresponds to the Stock 
A=cbind(as.matrix(c(103,103)), as.matrix(c(90,130))) 
# Add dimension names 
dimnames(A)<-list(c("Payoff_T_d","Payoff_T_u"),c("Bond","Stock")) 
A 
```

```{r}
# When using ggplot2, data are represented in data frames 
is.data.frame(A) 
```

```{r}
Atranspose.df<-data.frame(t(A)) 
# Add Asset_name as a field in Atranspose.df 
Atranspose.df$Asset_name<-dimnames(Atranspose.df)[[1]] 
Atranspose.df 
```

```{r}
library(ggplot2) 
# Figure AA: Plot payoff vectors of Two Assets at t=T 
#     Note: the function arrow() specifies the head of the arrows 
g1 <- ggplot(data=Atranspose.df,aes(x=Payoff_T_d,
                                    y=Payoff_T_u,
                                    label=Asset_name,
                                    color=Asset_name)) 
g1 + 
  geom_segment(aes(x=0,y=0,xend=Payoff_T_d,yend=Payoff_T_u),
               linewidth=3,
               arrow = arrow(length = unit(0.5, "cm"))) +
  geom_text(aes(label=Asset_name), hjust=0.,vjust=1., nudge_x=5, nudge_y=7) +
  xlim(-50,140) + 
  ylim(-50,140) + 
  ggtitle("Fig 1. Payoff_T Vectors of Two Assets") 
```

```{r}
# Create a data frame of the portfolio weights vectors
list.pi_S = seq(0, 1, 0.1)
df.pivecs.1 <- data.frame(pi_B = 1 - list.pi_S, pi_S = list.pi_S)
```

```{r}
# Create a data frame data.portfolios 
# with rows corresponding to portfolio cases which contain variables
#   pi_B  portfolio weights of B 
#   pi_S  portfolio weight of S 
#   Value_0  value/cost of portfolio at t=0 
#   Payoff_T_u  payoff of portfolio at t=T for omega = u 
#   Payoff_T_d  payoff of portfolio at t=T for omega = d 
#   Profit_T_u  profit of portfolio at t=T for omega = u 
#   Profit_T_d  profit of portfolio at t=T for omega = d

for (i in 1:NROW(df.pivecs.1)){
  pivec= as.numeric(df.pivecs.1[i,])
  Value_0=sum(A0*pivec) 
  Payoff_T_d=sum(A[1,]*pivec) 
  Payoff_T_u=sum(A[2,]*pivec) 
  Profit_T_d=Payoff_T_d-Value_0 
  Profit_T_u=Payoff_T_u-Value_0 

  new.data.portfolios <- data.frame(pi_B=pivec[1],
                                    pi_S=pivec[2], 
                                    Value_0=Value_0, 
                                    Payoff_T_d=Payoff_T_d, 
                                    Payoff_T_u=Payoff_T_u, 
                                    Profit_T_d=Profit_T_d, 
                                    Profit_T_u=Profit_T_u) 
  if (i==1){ 
    data.portfolios<-new.data.portfolios
  } else {
    data.portfolios<-rbind(data.portfolios, new.data.portfolios)
  }
}
data.portfolios.1=data.portfolios 

```

```{r}
# load library ggplot2 
library(ggplot2) 
data.portfolios.1$pi_S <-as.factor(data.portfolios.1$pi_S) 
# Figure 2A: Plot payoffs as vectors 
g1 <- ggplot(data=data.portfolios.1,aes(x=Payoff_T_d,y=Payoff_T_u)) 
g2 <- g1 + geom_segment(aes(x=0,
                            y=0,
                            xend=Payoff_T_d,
                            yend=Payoff_T_u,
                            col=pi_S),
                        linewidth=.5,
                        arrow = arrow(length = unit(0.2, "cm"))) +
  xlim(c(-140,140)) +
  ylim(c(-140,140)) 
g2 + ggtitle("Figure 2A. Payoff Vectors at t=T (as vectors)",
             subtitle="Constraints: weights Pi_B + weights Pi_S = 1") 
```

```{r}
# Figure 2B: Plot payoffs as points 
g1 <- ggplot(data=data.portfolios.1,aes(x=Payoff_T_d,y=Payoff_T_u)) 
g2 <- g1 +
  geom_point(aes(col=pi_S),size=1) +
  xlim(c(-140,140)) +
  ylim(c(-140,140)) 
g2 + ggtitle("Figure 2B. Payoff Vectors at t=T (as points)",
             subtitle="Constraints: weights Pi_B + weights Pi_S = 1") 
```

## Portfolios with Linear Constraints on Weights

Now, we consider different constraints on the portfolios by defining sets of portfolios satisfying certain constraints.

```{r}
# Write a function that creates a data frame with portfolio attributes 
# and receives portfolio weights in rows of an input data frame
  
fcn.data.portfolios <- function(df.pivecs, A0, A) {
  
  # Ensure inputs are in the correct numeric format
  # A0 should be a numeric vector (initial prices)
  # A should be a numeric matrix (rows = final state, columns = assets)
  A0 <- as.numeric(A0)
  A  <- as.matrix(A)
  
  # Convert portfolio weights data frame to matrix
  # Each row represents a portfolio (pi_B, pi_S)
  W <- as.matrix(df.pivecs)
  
  # Check dimension consistency
  # Number of assets must match
  if (ncol(W) != length(A0) || ncol(W) != ncol(A)) {
    stop("Dimension mismatch: check A0, A, and df.pivecs")
  }
  
  # Compute initial portfolio value for all portfolios
  # Each row: dot product between weights and initial prices
  Value_0 <- as.vector(W %*% A0)
  
  # Compute payoffs in each state
  # Row 1 of A = payoff in state d
  # Row 2 of A = payoff in state u
  Payoff_T_d <- as.vector(W %*% A[1, ])
  Payoff_T_u <- as.vector(W %*% A[2, ])
  
  # Compute profits in each state
  Profit_T_d <- Payoff_T_d - Value_0
  Profit_T_u <- Payoff_T_u - Value_0
  
  # Build final data frame with all portfolio characteristics
  data.portfolios <- data.frame(
    pi_B = W[, 1],
    pi_S = W[, 2],
    Value_0 = Value_0,
    Payoff_T_d = Payoff_T_d,
    Payoff_T_u = Payoff_T_u,
    Profit_T_d = Profit_T_d,
    Profit_T_u = Profit_T_u
  )
  
  # Return result
  return(data.portfolios)
}
```

## Portfolios that DON'T allow Short Selling

```{r}
# Create grid of all combinations
df.pivecs.2 <- expand.grid(
  pi_B = seq(0, 1, 0.1),
  pi_S = seq(0, 1, 0.1)
)

head(df.pivecs.2)
```

```{r}
data.portfolios.2<-fcn.data.portfolios(df.pivecs.2, A0, A) 
# Figure 3: Plot payoffs as points for data.portfolios.2 
g1 <- ggplot(data=data.portfolios.2,aes(x=Payoff_T_d,y=Payoff_T_u)) 
g2 <- g1 +
  geom_point(aes(col=Value_0),size=1) +
  xlim(c(-140,250)) +
  ylim(c(-140,250))
# Note: the xlim and ylim values are set manually 
subtitle <-"Constraints: 0 <= weights Pi_B , weights Pi_S <= 1"
g2 + ggtitle("Figure 3. Payoff Vectors at t=T (as points)",
             subtitle=subtitle) 
```

## Portfolios Allowing Shorting

Now, we consider portfolios that allow shorting of either the Bond B or the stock S.

```{r}
# Consider df.pivecs that let's pi_B and pi_S range between -2 and 2 
df.pivecs.3 <- expand.grid(
  pi_B = seq(-2, 2, 0.1),
  pi_S = seq(-2, 2, 0.1)
)

head(df.pivecs.3)
```

```{r}
data.portfolios.3<-fcn.data.portfolios(df.pivecs.3, A0, A) 
# Figure 4: Plot payoffs as points for data.portfolios.3 
g1<-ggplot(data=data.portfolios.3,aes(x=Payoff_T_d,y=Payoff_T_u))

g2<- g1 +
  geom_point(aes(col=Value_0),size=1) +
  scale_color_gradient2(
    low = "red",     # valores negativos
    mid = "white",    # valor 0
    high = "blue",     # valores positivos
    midpoint = 0
  ) + 
  xlim(c(-500,500)) +
  ylim(c(-500,500))

g2 + ggtitle("Figure 4. Payoff Vectors at t=T (as points)",
             subtitle="Constraints: -2 <= weights Pi_B , weights Pi_S <= +2") 

```

Note that the zero-cost portfolios (in white) do not admit positive payoffs in both states (u and d); i.e., **there is no arbitrage.**
