---
title: "Regression Analysis - MIT 18.642"
editor: visual
execute:
  echo: true

format:
  pdf:
    output-file: "mit18_642_f24_lec06_1_2 - Regression analysis"
    toc: true
    number-sections: true
    
  html:
    output-file: "mit18_642_f24_lec06_1_2 - Regression analysis"
    embed-resources: true
    image-format: "png"
    toc: true
    toc-location: "left-body"
    number-sections: true
---

\newpage

# Lecture 6 - MIT 18.642- Regression Analysis

```{r, warning=FALSE, message=FALSE}
# Load packages into R session (and install if necessary)

# require() returns TRUE if the package is successfully loaded
# and FALSE if it is not found.
if (!require("faraway")) install.packages("faraway")
if (!require("tidyverse")) install.packages("tidyverse")
library(faraway)
library(tidyverse)
library(car)
library(stats)
library(MASS) # For studres()
```

## Load data set Prostate

```{r}
# Load the dataset into the environment 
# Note: data() returns the dataset name,
# the actual values are stored in the 'prostate' dataframe
dataset_name <- data(prostate) 
print(dataset_name)

# Display the first 3 observations of the prostate dataframe
print(head(prostate, 3))

```

\newpage

## Variables:

| Variable | Description |
|:---|:---|
| **lcavol** | log cancer volume |
| **lweight** | log prostate weight |
| **age** | in years |
| **lbph** | log of the amount of benign prostatic hyperplasia |
| **svi** | seminal vesicle invasion |
| **lcp** | log of capsular penetration |
| **gleason** | a numeric vector |
| **pgg45** | percent of Gleason score 4 or 5 |
| **lpsa** | response (log prostate specific antigen) |

: Dataset Prostate (Library: faraway) {#tbl-prostate}

```{r}
round(apply(prostate,2,summary),digits=1)
```

```{r}
round(cor(prostate[,1:8]),digits=1) 
```

```{r}
pairs( prostate[,1:9], col="blue" ) 
```

## Regression on Original Covariates

```{r}

# Set a seed to ensure reproducibility
set.seed(1) 

# Create a data frame with a single column 
# containing the row indices of the prostate dataset
prostate_id <- data.frame(irow = 1:nrow(prostate)) 

# Randomly select 75% of the indices to be used for the training set
train_id <- slice_sample(prostate_id, prop = .75) 

# Subset the original dataset to create the training set
train <- prostate[train_id$irow, ] 

# Create the test set by excluding training indices (using the negative sign)
test <- prostate[-train_id$irow, ] 

# Fit a linear regression model using all available predictors 
# and summarize the results
linear_regression <- lm(lpsa ~ ., data = train)
summary(linear_regression)
```

## Regression on Standardized Covariates

```{r}
# Create a copy of the training set to avoid modifying the original data
train0 <- train 

# Center and scale the first 8 predictor variables (mean=0, SD=1)
train0[, 1:8] <- scale(train[, 1:8]) 

# Fit the linear model on scaled data and display the summary results
summary(lm(lpsa ~ ., data = train0))
```

## R-squared

```{r}
fit <- lm(lpsa~., data=train0)
fit.summary <- summary(fit) 
names(fit.summary)
```

```{r}
plot(fit$fitted.values,
     train0$lpsa,
     xlab="Fitted Value",
     ylab="Observed Value (lpsa)", 
     main=paste0(
       c("Prostate Model",
         "\nR-Squared =",
         as.character(round(fit.summary$r.squared, digits=2))
         ),
       collapse = ""
     )
) 
abline(lm(train0$lpsa ~ fit$fitted.values),col="blue",lwd=3) 
```

# Package MASS

## studres() in Package MASS 

```{r}
# Compute studentized residuals for the model
studentized_residuals <- studres(fit)
# Plot histogram of studentized residuals with 25 bins
hist(studentized_residuals, breaks = 25)
```

# Package car

## qqPlot() in Package car

```{r}
# Create a Quantile-Quantile plot to check for normality
Extreme_Residuals <- qqPlot(fit, col = "blue", col.lines = "red") 

```

The `qqPlot` function returns a **named integer vector** where the **values** represent the row positions of the most influential cases, and the **names** represent their corresponding row labels in the data set

```{r}
# Construct the message using the positions in the training set
message <- paste0("The QQ-plot identifies the observations at training set",
                  " positions:\n [",
                  paste(as.numeric(Extreme_Residuals), collapse = ", "),
                  "]\n as potential outliers.")

# Print the final result
cat(message)
```

```{r}
# Option A: Using the labels (The safest way)
outlier_data <- train0[names(Extreme_Residuals), ]
print(outlier_data)

# Option B: Using the values (Position in the training set)
outlier_data_alt <- train0[Extreme_Residuals, ]
print(outlier_data_alt)
```

## influencePlot() in Package car

Unlike `qqPlot` (which only identifies points with extreme residuals), `influencePlot` is more rigorous because it looks for observations that meet any of the following conditions:

-   **High Studentized Residuals (Outliers):** Observations where the model fails significantly to predict the response.

-   **High Leverage:** Points with extreme values in the predictor variables ($X$) that have the potential to pull the regression line toward them.

-   **High Global Impact (Cook's Distance):** Observations that, if removed, would substantially change the model's coefficients.

```{r}
# Store the influence diagnostics in a data frame
influence_points <- influencePlot(lm(lpsa ~ ., data = train0))
```

```{r}
# Display the coordinates and metrics of the most influential cases
print(influence_points)

```

```{r}
# To access the specific row names of these influential points
influential_names <- rownames(influence_points)
influential_names
```

## residualPlots() in Package car

```{r}
residualPlots(lm(lpsa~., data=train0),tests=TRUE,plot=FALSE) 
```

```{r}
residualPlots(lm(lpsa~., data=train0),tests=FALSE,plot=TRUE)
```

## leveragePlots() in Package car

```{r}
leveragePlots(lm(lpsa~., data=train0)) 
```

# Package STATS

## plot.lm() in STATS

```{r}
# The R function $plot.lm()$ generates a 2x2 display of plots 
oldpar=par(no.readonly=TRUE) 
layout(matrix(c(1,2,3,4),2,2)) # optional 4 graphs/page 
plot(fit)
par(oldpar,no.readonly=TRUE)  

```

